AdaptiveConformal: An `R` Package for Adaptive Conformal Inference

### 2023-10-09 paper.qmd
Updated paper draft

---
title: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
subtitle: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
author:
  - name: Herbert Susmann
    corresponding: true
    email: herbert.susmann@dauphine.psl.eu
    url: https://herbsusmann.com
    orcid: 0000-0002-3540-8255
    #affiliations:
    #  - name: CNRS, Université Paris Dauphine - PSL
    #    url: https://cnrs.fr
    #  - name: PR[AI]RIE Institute
    #    url: https://prairie-institute.fr/
  - name: Margaux Zaffran
    email: margaux.zaffran@inria.fr
    url: https://mzaffran.github.io/
    orcid: 0000-0001-7560-8916
    #affiliations:
    #  - name: INRIA
    #    url: https://inria.fr
    #  - name: CMAP, Ecole Polytechnique
    #    url: https://inria.fr
    #  - name: EDF R\&D
    #    url: https://inria.fr
  - name: Antoine Chambaz
    email: antoine.chambaz@u-paris.fr
    url: https://helios2.mi.parisdescartes.fr/~chambaz/
    orcid: 0000-0002-5592-6471
    #affiliations:
    #  - name: Université Paris Cité
    #    department: MAP5
    #    url: https://map5.mi.parisdescartes.fr/
  - name: Julie Josse
    email: julie.josse@inria.fr
    url: http://juliejosse.com/
    orcid: 0000-0001-9547-891X
    #affiliations:
    #  - name: INRIA Sophia-Antipolis
    #    url: https://www.inria.fr/fr/centre-inria-universite-cote-azur
date: last-modified
date-modified: last-modified
description: |
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
abstract: >+
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
keywords: [Conformal inference, Adaptive conformal inference, time series, R]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: computorg
repo: "template-computo-r"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: 
    code-fold: true
  computo-pdf: 
    include-in-header:
      - text: |
          \usepackage{stmaryrd}
          \usepackage{xfrac}
---

# Introduction
Conformal Inference (CI) is a family of methods for generating finite sample prediction intervals around point predictions when data are exchangeable [@shafer2008conformal; @angelopoulos2022gentle]. The input point predictions can be derived from any prediction method, making CI a powerful tool for augmenting black-box prediction algorithms with prediction intervals. Classical CI methods are able to yield marginally valid intervals with only the assumption that the joint distribution of the data does not change based on the order of the observations (that is, they are exchangeable). However, in many real-world settings data are not exchangeable: for example, time series data usually cannot be assumed to be exchangeable due to temporal dependence. A recent line of research examines the problem of generating prediction intervals for observations that are observed online (that is, one at a time) and for which exchangeability is not assumed to hold [@gibbs2021adaptive; @zaffran2022agaci; @gibbs2022faci; @bhatnagar2023saocp]. The methods from this literature, which we refer to generally as _Adaptive Conformal Inference_ (ACI) algorithms, work by adaptively adjusting the width of the generated prediction intervals in response to the observed data.

Informally, suppose a sequence of outcomes $y_t \in \mathbb{R}$, $t = 1, \dots, T$ are observed one at a time. Before seeing each observation, we have at our disposal a point prediction $\hat{\mu}_t \in \mathbb{R}$, which can be generated by any method. Our goal is to find an algorithm for producing prediction intervals $[\ell_t, u_t]$, $\ell_t \leq u_t$ such that, in the long run, the observations $y_t$ fall within the corresponding prediction intervals roughly $\alpha \times 100\%$ of the time: that is, $\lim_{T \to \infty} \sfrac{1}{T} \sum_{t=1}^T \mathbb{I}\{ y_t \in [\ell_t, u_t] \} = \alpha$. The original ACI algorithm [@gibbs2021adaptive] is based on a simple idea: if the previous prediction interval at time $(t-1)$ did not cover the true observation, then the next prediction interval at time $t$ is made slightly wider. Conversely, if the previous prediction interval did include the observation, then the next prediction interval is made slightly narrower. It can be shown that this procedure yields prediction intervals that in the long run cover the true observations the desired proportion of the time.

The main tuning parameter of the original ACI algorithm is a learning rate that controls how fast prediction interval width changes. If the learning rate is too low, then the prediction intervals will not be able to adapt fast enough to shifts in the data generating distribution; if it is too large, then the intervals will oscillate widely. The critical dependence of the original ACI algorithm on proper choice of its learning rate spurred subsequent research into meta-algorithms that learn the correct learning rate (or an analogue thereof) in various ways, typically drawing on approaches from the online learning literature. In this paper, we present four such algorithms: Aggregated ACI [AgACI, @zaffran2022agaci], Fully Adaptive Conformal Inference [FACI, @gibbs2022faci], Scale-Free Online Gradient Descent [SF-OGD, @bhatnagar2023saocp], and Strongly Adaptive Online Conformal Prediction [SAOCP, @bhatnagar2023saocp].

Our primary practical contribution is an implementation of each algorithm in an open source `R` package, `AdaptiveConformal`, which is available at [https://github.com/herbps10/AdaptiveConformal](https://github.com/herbps10/AdaptiveConformal). The package also includes routines for visualization and summary of the prediction intervals. We note that Python versions of several algorithms were also made available by @zaffran2022agaci and @bhatnagar2023saocp, but to our knowledge this is the first `R` package to implement them.

The rest of the paper unfolds as follows. In @sec-theory, we present a unified theoretical framework for analyzing the ACI algorithms based on the online learning paradigm. In @sec-algorithms we provide descriptions of each algorithm along with their known theoretical properties. In @sec-simulations we compare the performance of the algorithms in several simulation studies. @sec-case-study gives a case study of generating prediction intervals for influenza incidence in the United States based on black-box point forecasts. Finally, @sec-discussion provides a discussion and ideas for future research in this rapidly expanding field.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(future)
library(furrr)
library(progressr)

# Contents of helpers.R included in the appendix
source("helpers.R")
```

@@

We consider an online learning scenario in which we gain access to a sequence of observations $(y_t)_{t \geq 1}$ one at a time (see @cesabianchi2006games for an comprehensive account of online learning theory). Fix $\alpha \in (0, 1)$ to be the target empirical coverage of the prediction intervals. The goal is to output at time $t$ a prediction interval for the unseen observation $y_{t}$, with the prediction interval generated by an _interval construction function_ $\widehat{C}_{t+1}$. Formally, let $\widehat{C}_t$ be a function that takes as input a parameter $\theta_t \in \mathbb{R}$ and outputs a closed prediction interval $[\ell_t, u_t]$. The interval construction function must be nested: if $\theta^\prime > \theta$, then $\widehat{C}_t(\theta) \subseteq \widehat{C}_t(\theta^\prime)$. In words, larger values of $\theta$ imply wider prediction intervals. The interval constructor is indexed by $t$ to emphasize that it may use other information at each time point, such as a point prediction $\hat{\mu}_t \in \mathbb{R}$. We make no restrictions on how this external information is generated.

Define $r_t := \inf\{\theta \in \mathbb{R} : \mathbb{I}(y_t \in \widehat{C}_t(\theta)) \}$ to be the _radius_ at time $t$. The radius is the smallest possible $\theta$ such that the prediction interval covers the observation $y_t$. A key assumption for the theoretical analysis of several of the algorithms is that the radii are bounded: that is, there exists a $D > 0$ such that $r_t < D$ for all $t$.

Next, we describe two possible definitions of the interval construction function.

## Linear Intervals
A simple method for forming the prediction intervals is to use the parameter $\theta_t$ to directly define the width of the interval. Suppose that at each time $t$ we have access to a point prediction $\hat{\mu}_t \in \mathbb{R}$. Then we can form a symmetric prediction interval around the point estimate using
$$
\begin{aligned}
  \theta \mapsto \widehat{C}_t(\theta) := [\hat{\mu}_t - \theta, \hat{\mu}_t + \theta].
\end{aligned}
$$
We refer to this as the _linear interval constructor_. Note that in this case, the radius is simply the absolute residual $r_t = |\hat{\mu}_t - y_t|$. We focus primarily on this interval constructor in this paper, and it is the default used in the `AdaptiveConformal` `R` package. _Alternatively, we could also include results for all the algorithms run with conformal interval scores, perhaps in appendix_

## Conformal Intervals
The original ACI paper proposed constructing intervals based on the previously observed residuals [@gibbs2021adaptive]. Let $S : \mathbb{R}^2 \to \mathbb{R}$ be a function called a _conformity score_. A popular choice of conformity score is the absolute residual: $(\mu, y) \mapsto  S(\mu, y):= |\mu - y|$.  Let $s_t := S(\hat{\mu}_t, y_t)$ be the conformity score of the $t$th-observation. The conformal interval construction function is then given by
$$
\begin{aligned}
  \widehat{C}_t(\theta_t) := [\hat{\mu}_t - \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \}), \hat{\mu}_t + \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \})]
\end{aligned}
$$
where $\mathrm{Quantile}(\theta, A)$ denotes the empirical $\theta$-quantile of the elements in the set $A$. Note that $\widehat{C}_t$ is indeed nested in $\theta_t$ because the Quantile function is non-decreasing in $\theta$.

The `AdaptiveConformal` package takes the absolute residual as the default conformity score, although the user may also specify any custom conformity score by supplying it as an `R` function.

@@

## Assessing ACI algorithms
There are two different perspectives we can take in measuring the quality of an ACI algorithm that generates a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$. First, we could look at how close the empirical coverage of the generated prediction intervals is to the desired coverage level $\alpha$. Formally, define the empirical coverage as the proportion of observations that fell within the corresponding prediction interval: $\mathrm{EmpCov}(T) := \frac{1}{T} \sum_{t=1}^T (1 - \mathrm{err}_t)$. The coverage error is then given by
$$
\begin{aligned}
  \mathrm{CovErr}(T) := \mathrm{EmpCov}(T) - \alpha.
\end{aligned}
$$
The second perspective is to look at how well the algorithm controls the incurred pinball losses. Following the classical framework from the online learning literature, we define the _regret_ as the difference between the cumulative loss yielded by a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$ versus the cumulative loss of the best possible fixed choice:
$$
\begin{aligned}
  \mathrm{Reg}(T) := \sum_{t=1}^T L^\alpha(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=1}^T L^\alpha(\theta^*, r_t).
\end{aligned}
$$
In settings of distribution shift, it may not be appropriate to compare the cumulative loss of an algorithm to a fixed competitor. As such, stronger notions of regret have been defined. The _strongly adaptive regret_ is the largest regret over any subperiod of length $k \in \llbracket T \rrbracket$:
$$
\begin{aligned}
  \mathrm{SAReg}(T, k) := \max_{[\tau, \tau + k - 1] \subseteq \llbracket T \rrbracket} \left( \sum_{t=\tau}^{\tau + k - 1} L^{\alpha}(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=\tau}^{\tau + k - 1} L^\alpha(\theta^*, r_t) \right).
\end{aligned}
$$
Both ways of evaluating ACI methods are important because targeting only one or the other can lead to algorithms that yield prediction intervals that are not practically useful. As a simple pathological example of only targeting the coverage error, suppose we wish to generate $\alpha = 50\%$ prediction intervals. We could choose to alternate $\theta$ between 0 and $\infty$, such that $\mathrm{err}_t$ alternates between 0 and 1. The empirical coverage would then trivially converge to the desired level of 50\%. However, the same algorithm would yield infinite regret (see @bhatnagar2023saocp for a more in-depth example of an scenario in which coverage is optimal but the regret grows linearly). On the other hand, an algorithm that has arbitrarily small regret may not yield good empirical coverage. Suppose the observations and point predictions are constant: $y_t = 1$ and $\hat{\mu}_t = 0$ for all $t \geq 1$. Consider a simple class of algorithms that outputs constantly $\theta_t = \theta'$ for some $\theta' < 1$. With the linear interval construction function, the prediction intervals are then $\widehat{C}_t(\theta_t) = [-\theta', \theta']$. The regret is given by $\mathrm{Reg}(T) = 2T\alpha(1-\theta')$, which approaches zero as $\theta'$ approaches 1. The empirical coverage is, however, always zero. In other words, the regret can be arbitrarily close to zero while at the same time the empirical coverage does not approach the desired level.

These simple examples illustrate that, unfortunately, bounds on the coverage error and bounds on the regret are not in general interchangeable. It is possible, however, to show equivalencies by either (1) making distributional assumptions on the data or (2) using additional information about how the algorithm produces the sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$.

It may also be informative to summarize a set of prediction intervals in ways beyond their coverage error or their regret. A common metric for prediction intervals is the _mean interval width_:
$$
\begin{aligned}
  \mathrm{MeanWidth}(T) := \frac{1}{T} \sum_{t=1}^T w_t,
\end{aligned}
$$
where $w_t := u_t - \ell_t$ is the interval width at time $t$. Finally, we introduce a metric that is intended to capture pathological behavior that can arise with ACI algorithms where the prediction intervals oscillate between being extremely narrow and extremely wide. Define the _path length_ of prediction intervals generated by an ACI algorithm as
$$
\begin{aligned}
  \mathrm{PathLength}(T) := \sum_{t=2}^T |w_t - w_{t-1}|.
\end{aligned}
$$
A high path length indicates that the prediction intervals were variable over time, and a low path length indicates the prediction intervals were stable.

# Algorithms {#sec-algorithms}
In this section we describe five ACI algorithms and summarize the theoretical guarantees that have been proven for each one. We have harmonized the algorithm descriptions to use a common structure and notation, in some cases departing from their original descriptions. As a quick reference, the algorithms are summarized in @tbl-aci.

| Algorithm                                             | Tuning Parameters                                           | Citation            |
| ------------------------------------------------------|-------------------------------------------------------------|---------------------|
| Adaptive Conformal Inference (ACI)                    | Learning rate $\gamma$                                      | @gibbs2021adaptive  |
| Aggregated Adaptive Conformal Inference (AgACI)       | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | @zaffran2022agaci   |
| Fully Adaptive Conformal Inference (FACI)             | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | @gibbs2022faci      |
| Scale-Free Online Gradient Descent (SF-OGD)           | Learning rate $\gamma$ or maximum radius $D$                | @bhatnagar2023saocp |
| Strongly Adaptive Online Conformal Prediction (SAOCP) | Learning rate $\gamma$, lifetime multiplier $g$                 | @bhatnagar2023saocp |

@@

As a simple running example to illustrate each algorithm, we simulate ndependently $T = 500$ values $y_1, \dots, y_T$ following
$$
y_t \sim N(0, 0.2^2), \quad t \in \llbracket T \rrbracket.
$$
For demonstration purposes we assume to have access to unbiased predictions $\hat{\mu}_t = 0$ for all $t \in \llbracket T \rrbracket$. Throughout we set a target empirical coverage of $\alpha = 0.8$.

@@

::: {#fig-algo-aci fig-align=left fig-pos=H}
```{.pseudocode}
\begin{algorithmic}
\State \textbf{Input:} starting value $\theta_1$, learning rate $\gamma > 0$.
\For{$t = 1, 2, \dots, T$}
  \State \textbf{Output} prediction interval $\widehat{C}_t(\theta_t)$.
  \State Observe $y_t$.
  \State Evaluate $\mathrm{err}_t = \mathbb{I}[y_t \not\in \widehat{C}_t(\theta_t)]$.
  \State Update $\theta_{t+1} = \theta_t + \gamma (\mathrm{err}_t - (1 - \alpha))$.
\EndFor
\end{algorithmic}
```
Adaptive Conformal Inference algorithm [@gibbs2021adaptive].
:::

>>>>

---
title: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
subtitle: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
author:
  - name: Herbert Susmann
    corresponding: true
    email: herbert.susmann@dauphine.psl.eu
    url: https://herbsusmann.com
    orcid: 0000-0002-3540-8255
    affiliations:
      - name: CEREMADE (UMR 7534), Université Paris-Dauphine PSL, Place du Maréchal de Lattre de Tassigny, Paris, 75016, France
        url: https://www.ceremade.dauphine.fr/
  - name: Antoine Chambaz
    email: antoine.chambaz@u-paris.fr
    url: https://helios2.mi.parisdescartes.fr/~chambaz/
    orcid: 0000-0002-5592-6471
    affiliations:
      - name: MAP5 (UMR 8145), Université Paris Cité
        department: 
        url: https://map5.mi.parisdescartes.fr/
  - name: Julie Josse
    email: julie.josse@inria.fr
    url: http://juliejosse.com/
    orcid: 0000-0001-9547-891X
    affiliations:
      - name: Inria PreMeDICaL team, Université de Montpellier
        url: https://team.inria.fr/premedical/
date: last-modified
date-modified: last-modified
description: |
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
abstract: >+
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
keywords: [Conformal inference, Adaptive conformal inference, time series, R]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: computorg
repo: "template-computo-r"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: 
    code-fold: true
  computo-pdf: 
    include-in-header:
      - text: |
          \usepackage{stmaryrd}
          \usepackage{xfrac}
---

# Introduction
Conformal Inference (CI) is a family of methods for generating finite sample prediction intervals around point predictions when data are exchangeable [@vovk2005; @shafer2008conformal; @angelopoulos2022gentle]. The input point predictions can be derived from any prediction method, making CI a powerful tool for augmenting black-box prediction algorithms with prediction intervals. Classical CI methods are able to yield marginally valid intervals with only the assumption that the joint distribution of the data does not change based on the order of the observations (that is, they are exchangeable). However, in many real-world settings data are not exchangeable: for example, time series data usually cannot be assumed to be exchangeable due to temporal dependence. A recent line of research examines the problem of generating prediction intervals for observations that are observed online (that is, one at a time) and for which exchangeability is not assumed to hold [@gibbs2021adaptive; @zaffran2022agaci; @gibbs2022faci; @bhatnagar2023saocp]. The methods from this literature, which we refer to generally as _Adaptive Conformal Inference_ (ACI) algorithms, work by adaptively adjusting the width of the generated prediction intervals in response to the observed data.

Informally, suppose a sequence of outcomes $y_t \in \mathbb{R}$, $t = 1, \dots, T$ are observed one at a time. Before seeing each observation, we have at our disposal a point prediction $\hat{\mu}_t \in \mathbb{R}$ that can be generated by any method. Our goal is to find an algorithm for producing prediction intervals $[\ell_t, u_t]$, $\ell_t \leq u_t$ such that, in the long run, the observations $y_t$ fall within the corresponding prediction intervals roughly $\alpha \times 100\%$ of the time: that is, $\lim_{T \to \infty} \sfrac{1}{T} \sum_{t=1}^T \mathbb{I}\{ y_t \in [\ell_t, u_t] \} = \alpha$. The original ACI algorithm [@gibbs2021adaptive] is based on a simple idea: if the previous prediction interval at time $(t-1)$ did not cover the true observation, then the next prediction interval at time $t$ is made slightly wider. Conversely, if the previous prediction interval did include the observation, then the next prediction interval is made slightly narrower. It can be shown that this procedure yields prediction intervals that in the long run cover the true observations the desired proportion of the time.

The main tuning parameter of the original ACI algorithm is a learning rate that controls how fast prediction interval width changes. If the learning rate is too low, then the prediction intervals will not be able to adapt fast enough to shifts in the data generating distribution; if it is too large, then the intervals will oscillate widely. The critical dependence of the original ACI algorithm on proper choice of its learning rate spurred subsequent research into meta-algorithms that learn the correct learning rate (or an analogue thereof) in various ways, typically drawing on approaches from the online learning literature. In this paper, we present four such algorithms: Aggregated ACI [AgACI, @zaffran2022agaci], Fully Adaptive Conformal Inference [FACI, @gibbs2022faci], Scale-Free Online Gradient Descent [SF-OGD, @bhatnagar2023saocp], and Strongly Adaptive Online Conformal Prediction [SAOCP, @bhatnagar2023saocp]. We note that the adaption of conformal inference techniques is an active area of research and the algorithms we focus on in this work are not exhaustive; see among others @feldman2023achieving, @bastani2022practical, @xu2021enbpi, and @xu2023spci.

Our primary practical contribution is an implementation of each algorithm in an open source `R` package, `AdaptiveConformal`, which is available at [https://github.com/herbps10/AdaptiveConformal](https://github.com/herbps10/AdaptiveConformal). The package also includes routines for visualization and summary of the prediction intervals. We note that Python versions of several algorithms were also made available by @zaffran2022agaci and @bhatnagar2023saocp, but to our knowledge this is the first package implementing them in `R`. In addition, several `R` packages exist for conformal inference in other contexts, including `conformalInference` focusing on regression [@tibshirani2019ci], `conformalInference.fd`, with methods for functional responses [@diquigiovanni2022fd], and `cfcausal` for causal inference related functionals [@lei2020cfcausal]. Our second practical contribution is to compare the performance of the algorithms in simulation studies and in a case study generating prediction intervals for influenza incidence in the United States based on black-box point forecasts.

The rest of the paper unfolds as follows. In @sec-theory, we present a unified theoretical framework for analyzing the ACI algorithms based on the online learning paradigm. In @sec-algorithms we provide descriptions of each algorithm along with their known theoretical properties. In @sec-simulations we compare the performance of the algorithms in several simulation studies. @sec-case-study gives a case study based on forecasting influenza in the United States. Finally, @sec-discussion provides a discussion and ideas for future research in this rapidly expanding field.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(patchwork)
library(future)
library(furrr)
library(progressr)

knitr::opts_chunk$set(echo = FALSE)

source("helpers.R")
```

@@

We consider an online learning scenario in which we gain access to a sequence of observations $(y_t)_{t \geq 1}$ one at a time (see @cesabianchi2006games for an comprehensive account of online learning theory). Fix $\alpha \in (0, 1)$ to be the target empirical coverage of the prediction intervals. The goal is to output at time $t$ a prediction interval for the unseen observation $y_{t}$, with the prediction interval generated by an _interval construction function_ $\widehat{C}_{t}$. Formally, let $\widehat{C}_t$ be a function that takes as input a parameter $\theta_t \in \mathbb{R}$ and outputs a closed prediction interval $[\ell_t, u_t]$. The interval construction function must be nested: if $\theta^\prime > \theta$, then $\widehat{C}_t(\theta) \subseteq \widehat{C}_t(\theta^\prime)$. In words, larger values of $\theta$ imply wider prediction intervals. The interval constructor is indexed by $t$ to emphasize that it may use other information at each time point, such as a point prediction $\hat{\mu}_t \in \mathbb{R}$. We make no restrictions on how this external information is generated.

Define $r_t := \inf\{\theta \in \mathbb{R} : \mathbb{I}(y_t \in \widehat{C}_t(\theta)) \}$ to be the _radius_ at time $t$. The radius is the smallest possible $\theta$ such that the prediction interval covers the observation $y_t$. A key assumption for the theoretical analysis of several of the algorithms is that the radii are bounded:

**Assumption**: there exists a $D > 0$ such that $r_t < D$ for all $t$.

Next, we describe two existing definitions of interval construction functions.

## Linear Intervals
A simple method for forming the prediction intervals is to use the parameter $\theta_t$ to directly define the width of the interval. Suppose that at each time $t$ we have access to a point prediction $\hat{\mu}_t \in \mathbb{R}$. Then we can form a symmetric prediction interval around the point estimate using
$$
\begin{aligned}
  \theta \mapsto \widehat{C}_t(\theta) := [\hat{\mu}_t - \theta, \hat{\mu}_t + \theta].
\end{aligned}
$$
We refer to this as the _linear interval constructor_. Note that in this case, the radius is simply the absolute residual $r_t = |\hat{\mu}_t - y_t|$.

## Quantile Intervals
The original ACI paper proposed constructing intervals based on the previously observed residuals [@gibbs2021adaptive]. Let $S : \mathbb{R}^2 \to \mathbb{R}$ be a function called a _nonconformity score_. A popular choice of nonconformity score is the absolute residual: $(\mu, y) \mapsto  S(\mu, y):= |\mu - y|$.  Let $s_t := S(\hat{\mu}_t, y_t)$ be the nonconformity score of the $t$th-observation. The quantile interval construction function is then given by
$$
\begin{aligned}
  \widehat{C}_t(\theta_t) := [\hat{\mu}_t - \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \}), \hat{\mu}_t + \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \})]
\end{aligned}
$$
where $\mathrm{Quantile}(\theta, A)$ denotes the empirical $\theta$-quantile of the elements in the set $A$. Note that $\widehat{C}_t$ is indeed nested in $\theta_t$ because the Quantile function is non-decreasing in $\theta$.

Our proposed `AdaptiveConformal` package takes the absolute residual as the default nonconformity score, although the user may also specify any custom nonconformity score by supplying it as an `R` function.

@@

This iterative procedure is at the core of the online learning theoretical framework in which theoretical results have been derived.

## Assessing ACI algorithms
There are two different perspectives we can take in measuring the quality of an ACI algorithm that generates a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$. First, we could look at how close the empirical coverage of the generated prediction intervals is to the desired coverage level $\alpha$. Formally, define the empirical coverage as the proportion of observations that fell within the corresponding prediction interval: $\mathrm{EmpCov}(T) := \frac{1}{T} \sum_{t=1}^T (1 - \mathrm{err}_t)$. The coverage error is then given by
$$
\begin{aligned}
  \mathrm{CovErr}(T) := \mathrm{EmpCov}(T) - \alpha.
\end{aligned}
$$
The second perspective is to look at how well the algorithm controls the incurred pinball losses. Following the classical framework from the online learning literature, we define the _regret_ as the difference between the cumulative loss yielded by a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$ versus the cumulative loss of the best possible fixed choice:
$$
\begin{aligned}
  \mathrm{Reg}(T) := \sum_{t=1}^T L^\alpha(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=1}^T L^\alpha(\theta^*, r_t).
\end{aligned}
$$
In settings of distribution shift, it may not be appropriate to compare the cumulative loss of an algorithm to a fixed competitor. As such, stronger notions of regret have been defined. The _strongly adaptive regret_ is the largest regret over any subperiod of length $k \in \llbracket T \rrbracket$:
$$
\begin{aligned}
  \mathrm{SAReg}(T, m) := \max_{[\tau, \tau + m - 1] \subseteq \llbracket T \rrbracket} \left( \sum_{t=\tau}^{\tau + m - 1} L^{\alpha}(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=\tau}^{\tau + m - 1} L^\alpha(\theta^*, r_t) \right).
\end{aligned}
$$
Both ways of evaluating ACI methods are important because targeting only one or the other can lead to algorithms that yield prediction intervals that are not practically useful. As a simple pathological example of only targeting the coverage error, suppose we wish to generate $\alpha = 50\%$ prediction intervals. We could choose to alternate $\theta$ between 0 and $\infty$, such that $\mathrm{err}_t$ alternates between 0 and 1. The empirical coverage would then trivially converge to the desired level of 50\%. However, the same algorithm would yield infinite regret (see @bhatnagar2023saocp for a more in-depth example of an scenario in which coverage is optimal but the regret grows linearly). On the other hand, an algorithm that has arbitrarily small regret may not yield good empirical coverage. Suppose the observations and point predictions are constant: $y_t = 1$ and $\hat{\mu}_t = 0$ for all $t \geq 1$. Consider a simple class of algorithms that outputs constantly $\theta_t = \theta'$ for some $\theta' < 1$. With the linear interval construction function, the prediction intervals are then $\widehat{C}_t(\theta_t) = [-\theta', \theta']$. The regret is given by $\mathrm{Reg}(T) = 2T\alpha(1-\theta')$, which approaches zero as $\theta'$ approaches 1. The empirical coverage is, however, always zero. In other words, the regret can be arbitrarily close to zero while at the same time the empirical coverage does not approach the desired level.

These simple examples illustrate that, unfortunately, bounds on the coverage error and bounds on the regret are not in general interchangeable. It is possible, however, to show equivalencies by either (1) making distributional assumptions on the data or (2) using additional information about how the algorithm produces the sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$ [@bhatnagar2023saocp].

It may also be informative to summarize a set of prediction intervals in ways beyond their coverage error or their regret. A common metric for prediction intervals is the _mean interval width_:
$$
\begin{aligned}
  \mathrm{MeanWidth}(T) := \frac{1}{T} \sum_{t=1}^T w_t,
\end{aligned}
$$
where $w_t := u_t - \ell_t$ is the interval width at time $t$.

Finally, we introduce a metric that is intended to capture pathological behavior that can arise with ACI algorithms where the prediction intervals oscillate between being extremely narrow and extremely wide. Define the _path length_ of prediction intervals generated by an ACI algorithm as
$$
\begin{aligned}
  \mathrm{PathLength}(T) := \sum_{t=2}^T |w_t - w_{t-1}|.
\end{aligned}
$$
A high path length indicates that the prediction intervals were variable over time, and a low path length indicates the prediction intervals were stable.

# Algorithms {#sec-algorithms}

| Algorithm                                             | Tuning Parameters                                           | Original interval constructor | Citation            |
| ------------------------------------------------------|-------------------------------------------------------------|---------------------|
| Adaptive Conformal Inference (ACI)                    | Learning rate $\gamma$                                      | Quantile | @gibbs2021adaptive  |
| Aggregated Adaptive Conformal Inference (AgACI)       | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | Quantile | @zaffran2022agaci   |
| Fully Adaptive Conformal Inference (FACI)             | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | Quantile | @gibbs2022faci      |
| Scale-Free Online Gradient Descent (SF-OGD)           | Learning rate $\gamma$ or maximum radius $D$             | Linear | @bhatnagar2023saocp |
| Strongly Adaptive Online Conformal Prediction (SAOCP) | Learning rate $\gamma$, lifetime multiplier $g$          | Linear | @bhatnagar2023saocp |

@@

As a simple running example to illustrate each algorithm, we simulate independently $T = 500$ values $y_1, \dots, y_T$ following
$$
y_t \sim N(0, 0.2^2), \quad t \in \llbracket T \rrbracket.
$$
For demonstration purposes we assume we have access to unbiased predictions $\hat{\mu}_t = 0$ for all $t \in \llbracket T \rrbracket$. Throughout we set the target empirical coverage to $\alpha = 0.8$.

@@

```pseudocode
#| label: algo-aci
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true

\begin{algorithm}
\caption{Adaptive Conformal Inference}
\begin{algorithmic}
\State \textbf{Input:} starting value $\theta_1$, learning rate $\gamma > 0$.
\For{$t = 1, 2, \dots, T$}
  \State \textbf{Output:} prediction interval $\widehat{C}_t(\theta_t)$.
  \State Observe $y_t$.
  \State Evaluate $\mathrm{err}_t = \mathbb{I}[y_t \not\in \widehat{C}_t(\theta_t)]$.
  \State Update $\theta_{t+1} = \theta_t + \gamma (\mathrm{err}_t - (1 - \alpha))$.
\EndFor
\end{algorithmic}
\end{algorithm}
```
